---
title: "When and how to use container"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{When and how to use container}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
require(container)
knitr::opts_chunk$set(
  comment = "#",
    error = FALSE,
     tidy = FALSE,
    cache = FALSE,
 collapse = TRUE
)
options(width = 100L)
```

## When *not* to use container

Don't bother using the `container` framework if ...

* *speed* is of high importance.

The `container` package is written in pure R and was never designed or
optimized for speed. So if computation speed is critical for your application,
you rather should fall back to base R *list* or use packages that were 
optimized for performance,
such as the [collections](https://CRAN.R-project.org/package=collections)
package.

* your application is *simple*.

If the complexity of your application is low, it is probably not
worth adding the `container` dependency and, again, you are probably fine with
just using base R lists or vectors.


## When to use container
Consider using the `container` framework over base `list` if you are looking 
for ...

* lean but **safe** list operations in **critical** code, specifically to
  * recognize name clashes
  * prevent unintended NULLs
  * avoid access of undefined values
* **specialized** data structures ([Deque, Set, and Dict](deque-set-dict.html))
  with list properties.
* lists with [reference semantics](reference-semantics.html).

The following sections cover the first of the above points going through
all relevant access, replace and removal operations and showing the 
advantages of using the container package. For the other points,
please follow the respective links.

## Container operations
Since R has always been used as an interactive tool, base R list operations
are generous with respect to errors, which can lead to undetected errors.
In contrast, the container package provides additional functions for all basic 
operations that allow for fine control, avoid many pitfalls that can happen
with lists, and will usually result in more robust code.

### Add elements
New elements can be added "as usual" by concatenation or name.
```{r}
co <- container()
co[["x"]] <- 1
co <- c(co, 2)
co
```
In addition, the container package provides *add*, which 
allows to add new elements by name even if that name exists already.
```{r}
co = add(co, x = 3)  # same as c(co, container(x = 3))
co
```

### Replace elements
Basic replacement again can be done "as usual" by name or position.
```{r}
co[["x"]] <- 0
co[[2]] <- 12
co
```
In contrast to base *lists*, the container will not allow to add elements at
positions longer than the length of the object.
```{r, error = TRUE}
co[[4]] <- 4
```

If the name does not exist, the element is appended as known from base lists.
```{r}
co[["y"]] <- 4
co
```

If you want to make sure that something is replaced, *container* provides the
function *replace_at*, which will only replace elements if at names or
positions that exist. The following statements are all equal and show the 
different possibilities on how to use *replace_at*.
```{r}
replace_at(co, x = 10, y = 13)            # name = value pairs
replace_at(co, c("x", "y"),  c(10, 13))   # names followed by values 
replace_at(co, c(1, 4),      c(10, 13))   # positions followed by values 
replace_at(co, list(1, "y"), c(10, 13))   # mixed names/positions followed by values
```

Now see how invalid indices are signaled.
```{r, error = TRUE}
replace_at(co, z = 10)
replace_at(co, "z", 10)
replace_at(co, 5, 10)
```

If you instead want elements at new names to be added, set `.add = TRUE`.
Invalid positional indices are still signaled.
```{r, error = TRUE}
co = replace_at(co, z = 10, .add = TRUE)
co = replace_at(co, 7, 10, .add = TRUE)
co
```

It is also possible to *replace* elements by value, that is, you specify the 
value (not the index) that should be replaced. Let's replace the 12 by "foo" and
then 4 by `1:4`.
```{r}
co = replace(co, 12, "foo")
co
co = replace(co, 4, 1:4)
co
```

In an interactive R session you may want to apply the notation using curly
braces.
```{r}
co[[{1:4}]] <- 1:2
co
```

### Extract elements
First of all, standard extract operators apply as expected.
```{r}
co[[1]]
co[["x"]]
co[3:5]
co[c("x", "y", "z")]
```

Programmatically, the corresponding functions to select one or multiple
elements are named *at2* and *at*.
```{r}
at2(co, 1)
at2(co, "x")
at(co, 3:5)
at(co, c("x", "y", "z"))
```

As before you can specify mixed indices via lists.
```{r}
indices = list("x", 4, "z")
at(co, indices)
```

Again, accessing non-existent names or positions is signaled with an error.
```{r, error = TRUE}
at2(co, 10)
at2(co, "a")
at(co, 3:6)
at(co, c("x", "a"))
```

With base R lists non-existent indices usually yield `NULL`.
```{r}
l = list()
l[["a"]]
l[2:3]
```

If needed, the (less strict) *list* access can be mimicked with *peek_at* and
*peek_at2*.
```{r}
peek_at2(co, "a")
peek_at(co, 10, 11)
peek_at(co, 5:10)
```

As you see, one important difference is multiple access via *peek_at* will not
insert `NULL` values by default.
However, both functions in fact allow to specify the default value that is
returned if the index does not exist.
```{r}
peek_at2(co, "a", default = -1)

peek_at(co, "z", "a", .default = -1)
peek_at(co, 4:8, .default = NA)
```


### Remove elements
To remove elements in lists, they have to be replaced by `NULL`.
```{r}
l = list(a = 1)
l
l[["a"]] <- NULL
l
```

With the container package this is done differently, as replacing by `NULL`
will not delete the element but literally replace it by `NULL`.
```{r}
co[["x"]] <- NULL
co
```

Instead, elements can be deleted by index (*delete_at*) or value (*delete*)
as follows.
```{r}
delete_at(co, 1, "y", "z")
delete(co, NULL, 1:2, 10)   # same but remove by value
```

As before, invalid indices or missing values are signaled.
```{r, error = TRUE}
delete_at(co, "a")
delete_at(co, 10)

delete(co, 1:3)
```


If you need a less strict delete operation, use the *discard* functions, which
delete all valid indices/values and ignore the rest.
```{r}
discard_at(co, 1, "a")
discard_at(co, 1:100)

discard(co, NULL, 1:2, 1:3, 1:4)
```


### Combine containers
The *update* function is used to combine/merge two containers.
```{r}
c1 = container(1, b = 2)
c2 = container(   b = 0, c = 3)

update(c1, c2)
update(c2, c1)
```

With the container package this function is also provided for base R lists.
```{r}
l1 = list(1, b = 2)
l2 = list(   b = 0, c = 3)

update(l1, l2)
update(l2, l1)
```

pkgdown::build_article("code-development")










