---
title: "Get started"
author: "Roman Pahl"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction

The *container* package is designed to serve three main goals:

1. enhance base R *list* for interactive sessions
2. enhance base R *list* for code development
3. provide specialized data structures that are not available with base R

This vignette will cover the first point. For the second point see
vignette [Why and how container for code development](code-development.html).
Examples for specialized data structures are found in
[Manage parameter lists with dict](parameter-list.html)
and [Enhancing data.table with dict.table](dict.table.html).

```{r load-package}
library(container, warn.conflicts = FALSE)
co <- container(data = iris, f = Sepal.Length ~ Sepal.Width, n = 1:10)
```

**Print** nice and compact.
```{r}
co
```

**Access** elements by position or name or *both*
```{r}
co[1:2]
co[1, 3]
```
```{r}
co[["n"]]
```
```{r}
co[2, 3, "data"]
```

**Replace** elements by position or name or *value*
```{r}
co[[3]] <- 1:3
```
```{r}
co[["f"]] <- speed ~ dist
```
```{r}
co[[{iris}]] <- cars
co
```

**Update** names and elements with ease
```{r}
c1 = container(a = 1, b = 2)
c2 = container(a = 3, b = 4)
ref_rename(c1, "a", "A")
update(c1, c2)
```
