---
title: "Manage data columns with dict.table"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
    The dict.table as a mix of data.table and Dict extends the data.table by
    functions to enhance data column management. This vignette compares basic
    dplyr and dict.table data column operations and shows how both frameworks 
    can be easily combined.
vignette: >
  %\VignetteIndexEntry{Manage data columns with dict.table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
require(container)
require(dplyr)
library(microbenchmark)
library(ggplot2)
library(data.table)
library(tibble)


knitr::opts_chunk$set(
  comment = "#",
  prompt = F,
  tidy = FALSE,
  cache = FALSE,
  collapse = T,
  fig.width = 7
)

options(width = 100L)
```

## Motivation
The [dplyr](https://CRAN.R-project.org/package=dplyr) functions `select` and
`mutate` are widely used to manage data `data.frame` (or `tibble`) columns.
They cover a wide range of use cases and are applied in quick data exploration
as well as in data analysis pipelines.
On the other hand, when implementing critical code or building R packages,
developers may revert to base R to minimize errors and code dependencies. At
least, both `mutate` and `select` may require additional checking, for example,
to catch column name clashes.

The [container](https://cran.r-project.org/web/packages/container/index.html)
package in parts was developed to close this gap. With version 1.0.0,
it provides `dict.table`, which can be considered a
[data.table](https://CRAN.R-project.org/package=data.table) with
an extended set of functions to add, extract, remove and replace data
columns with minimal required additional checking, hopefully resulting in
lean and robust code.

This vignette compares basic `dplyr` and `dict.table` data column operations
and at the end shows that both frameworks can be easily combined.

## Column operations

To keep matters simple, we use a tiny data set.
```{r}
library(container)
library(dplyr)

data <- dict.table(x = c(0.2, 0.5), y = letters[1:2])
data
```


### Add
Let's add columns using `mutate`.
```{r}
data %>%
    mutate(ID = 1:2, z = 1)
```

For someone not familar with the [tidyverse](https://www.tidyverse.org/), this code block might read somewhat odd as the column is added and *not* mutated. To add a column via `dict.table` use `add`.
```{r}
data %>%
    add(ID = 1:2, z = 1)
```

The intend to *add* a column thus is stated more clearly. Next, instead of ID, let's add another numeric column `y`, which happens to "name-clash" with the already existing column.
```{r}
data %>%
    mutate(y = 1)
```

Of course, the initial y-column has been overwritten. While this was easy to see here, it may not if the data has a lot of columns or if column names are created dynamically during runtime. To catch this, usually some overhead is required.
```{r, error=TRUE}
if ("y" %in% colnames(data)) {
    stop("column y already exists")
} else {
    data %>%
        mutate(y = 1)
}
```

Let's see the `dict.table`-operation in comparison.
```{r, error=TRUE}
data %>%
    add(y = 1)
```
The name clash is caught by default and therefore requires no additional
checking.

### Modify
If the intend was indeed to overwrite the value, the `dict.table` function `replace_at` can be used.
```{r}
data %>%
    replace_at(y = 1)

# or programmatically
data %>%
    replace_at("y", 1)
```

As we saw above, if a column does not exist, `mutate` silently creates it for you. If this is not what you want, which means, you want to make sure something is overwritten, again, a workaround is needed.
```{r, error=TRUE}
if ("ID" %in% colnames(data)) {
    data %>%
        mutate(ID = 1:2)
} else {
    stop("column ID not in data.frame")
}
```

Once again, the workaround is already "built-in" in the `dict.table`-framework,
```{r, error=TRUE}
data %>%
    replace_at(ID = 1:2)
```
that is, `replace_at` expects the column to exist.

If we were to paraphrase the intend of the `mutate` function, it probably would be something
like *"Replace a column or, if it does not exist, add it."*.
As you may already have guessed, this can also be expressed within the `dict.table`-framework.

```{r}
data %>%
    replace_at(ID = 1:2, .add = TRUE)
```

### Extract

### Remove
A common [tidyverse](https://www.tidyverse.org/) approach to remove a column is based on
the `select` function. One corresponding `dict.table`-function is `delete`.
```{r}
data %>%
  select(-"y")

data %>%
    delete_at("y")
```

Let's see what happens if the column does not exist in the first place.
```{r, error=TRUE}
data %>%
    select(-"ID")

data %>%
    delete_at("ID")
```

So in this case, both frameworks will complain. Now assume we want the column to be removed if it exist but otherwise silently ignore the command, for example:
```{r}
if ("ID" %in% colnames(data)) {
    data %>%
        select(-"ID")
}
```

The `dict.table` provides a straight-forward solution via the `discard` function:
```{r}
data %>%
    discard_at("ID")
```

## Benchmark
To compare the performance of both frameworks, we benchmark some column
operations using the standard 'cars' data set.
As a hallmark reference we use [data.table](https://CRAN.R-project.org/package=data.table).


```{r}
library(microbenchmark)
library(ggplot2)
library(data.table)
library(tibble)

data = cars
head(cars)
```

For the benchmark, we add, replace and finally delete a column.
```{r benchmark1, warning = FALSE, message = FALSE, cache=TRUE}
bm <- microbenchmark(control = list(order="inorder"), times = 100,

    dict.table =
        as.dict.table(data) %>%
        add(time = .[["dist"]] / .[["speed"]]) %>%
        replace_at(dist = 0) %>%
        delete_at("speed"),

    `data.table[` =
        as.data.table(data)[
        ][, time := dist / speed
        ][, dist := 0
        ][, speed := NULL],

    dplyr =
        as_tibble(data) %>%
        mutate(time = dist / speed) %>%
        mutate(dist = 0) %>%
        select(-speed)
)
autoplot(bm) + theme_bw()
```

While `dict.table` and `data.table` performed nearly the same there is
some distance to `dplyr` (about 10x). Let's examine each operation in more detail.

```{r benchmark2, warning = FALSE, message = FALSE, cache=TRUE}
data = cars
bm <- microbenchmark(control = list(order="inorder"), times = 100,

    dit <- as.dict.table(data),
    dit <- add(dit, time = dit[["dist"]] / dit[["speed"]]),
    dit <- replace_at(dit, dist = 0),
    dit <- delete_at(dit, "speed"),

    dat <-  as.data.table(data),
    dat[, time := dist / speed],
    dat[, dist := 0],
    dat[, speed := NULL],

    tbl <- as_tibble(data),
    tbl <- mutate(tbl, time = dist / speed),
    tbl <- mutate(tbl, dist = 0),
    tbl <- select(tbl, -speed)
)
autoplot(bm) + theme_bw()
```

Apparently, the mutate and select operations are the slowest in comparison,
which for the most part should be a result of these functions providing
non-standard evaluation (NSE) and generally a wide range of ways to specify the
desired operation. Unsurprisingly such flexibility comes at a cost.

Since the
[data.table](https://CRAN.R-project.org/package=data.table) expressions also
involve NSE terms and some overhead, in this benchmark the `dict.table` performs
even best.
Having said that, of course, the [data.table](https://CRAN.R-project.org/package=data.table)
code can be further improved by avoiding the overhead and instead use reference semantics
via the `data.table` built-in `set` function.

```{r benchmark3, warning = FALSE, message = FALSE, cache=TRUE}
data = cars
bm <- microbenchmark(control = list(order="inorder"), times = 100,

    dict.table =
        as.dict.table(data) %>%
        add(time = dit[["dist"]] / dit[["speed"]]) %>%
        replace_at(dist = 0) %>%
        delete_at("speed"),

    ref_dict.table =
        as.dict.table(data) %>%
        ref_add(time = .[["dist"]] / .[["speed"]]) %>%
        ref_replace_at(dist = 0) %>%
        ref_delete_at("speed"),


    `data.table[` =
        as.data.table(data)[
        ][, time := dist / speed
        ][, dist := 0
        ][, speed := NULL],

    set_data.table =
        as.data.table(data) %>%
        set(j = "ID", value = .[["dist"]] / .[["speed"]]) %>%
        set(j = "dist", value = 0) %>%
        set(j = "speed", value = NULL)
)

autoplot(bm) + theme_bw()
```

This puts things back into perspective. We also provided a `dict.table`
version using reference semantic, which is also `built-in` and results in a
slight speed improvement over the standard version. As a result, `data.table`
remains the way to go when speed is key.

## Combine dplyr and dict.table
Since a `dict.table` is fully compatible with `dplyr` and `data.table`,
all of the presented frameworks can be easily combined in any order.

```{r}
res = data %>%
    as.dict.table %>%
    .[, time := dist / speed] %>%   # data.table
    replace_at(dist = 0) %>%        # container
    select(-speed)                  # dplyr
```


## Summary
In critical code, it is usually of high priority to avoid unintended data
column operations. For this, usually additional code is required to check
for the existence or absence of columns.
The `dict.table` framework provides a set of column operations with built-in
checking, thereby yielding safer and leaner code out of the box and
ultimately freeing the developer from writing some annoying checks over and
over again.

