---
title: "S3 methods overview and examples"
author: "Roman Pahl"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{S3 methods overview and examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
The `Container` class serves as the base class for `Deque`, `Set` and `Dict`, which inherit all methods from `Container`, except those that are overwritten. In addition, the `Container` and all its subclasses are iterable, that is, they provide a method returning an `Iterator` to iterate through the elements of the container object. The S3 interface provides constructor methods for the `Container` class as well as all its derived classes `Deque`, `Set` and `Dict`, which return objects of these respective classes. Then for each object, S3 methods and operators are provided to emulate the corresponding member methods.

The following table shows member methods divided by class. The top half contains all `Container`-related methods, each derived by the subclasses to the right unless there is a new entry in a sub-class column, meaning the method is overwritten by the subclass.
The bottom half contains methods unique to each subclass.

| Container                      | Deque               | Set              | Dict                         |
|--------------------------------|---------------------|------------------|------------------------------|
| `cont <- container()`          | `deq <- deque()`    | `s <- set()`     | `d <- dict()`                |
|                                |                     |                  |                              |
| `add(cont, elem)`              |                     | `add(s, elem)`   | `add(d, key, val),        d[key] <- val`|
| `clear(cont)`                  |                     |                  |                              |
| `discard(cont, elem, right=F)` |                     |                  | `discard(d, key)`            |
| `empty(cont)`                  |                     |                  |                              |
| `has(cont, elem)`              |                     |                  | `has(d, key)`                |
| `print(cont, list.len=10)`     |                     |                  |                              |
| `remove(cont, elem, right=F)`  |                     |                  | `remove(d, key)`             |
| `size(cont)`                   |                     |                  |                              |
| `type(cont)`                   |                     |                  |                              |
| `values(cont)`                 |                     |                  |                              |
|                                |                     |                  |                              |
|                                | `addleft(deq, elem)`| `s1 + s2`        | `getval(d, key), d[[key]]`   |
|                                | `count(deq, elem)`  | `s1 / s2`        | `keys(d)`                    |
|                                | `peek(deq)`         | `s1 - s2`        | `peek(d, key, default=NULL), d[key]` |
|                                | `peekleft(deq)`     | `s1 == s2`       | `pop(d, key)`                |
|                                | `pop(deq)`          | `s1 < s2`        | `popitem(d)`                 |
|                                | `popleft(deq)`      | `s1 > s2`        | `setval(d, key, val, add=F),        d[[key, add=F]] <- val` |
|                                | `reverse(deq)`      |                  | `sortkey(decr=FALSE)`        |
|                                | `rotate(deq, n=1L)` |                  | `update(d, other)`           |


Method descriptions are found in the respective online helps (see `?container`, `?deque`, `?set`, and `?dict`).

## Container
Objects created using the base `container` function are ready to be used. Examples of specialized objects using `deque`, `set`, and `dict`, follow below.
```{r lib-setup}
library(container)
```

```{r}
collection <- container()
empty(collection)
```

Since the created objects are still the same as when created via the `R6` interface, they always work both ways:
```{r}
size(collection)

collection$size()
```


#### add
By default, elements internally are stored in a basic `list` and therefore can be of any type.
```{r}
add(collection, 1)
add(collection, "A")
add(collection, data.frame(B=1, C=2))
type(collection)
```

The internal representation can always be retrieved directly using the `values` function.
```{r}
values(collection)
```

The container's print method presents the content more compact similar to `utils::str`
```{r}
print(collection)
```

If initialized with an R object, the type of the object is adopted to allow for efficient internal representations, if required.
```{r}
ints <- container(integer())
type(ints)
```

Initialization also works with vectors.
```{r}
ints <- container(1:10)
print(ints)

values(ints)

size(ints)
```

#### discard, remove
```{r}
has(ints, 11)

has(ints, 7)

discard(ints, 7)
has(ints, 7)

remove(ints, 8)
values(ints)
```

Using `remove` on non-existent elements throws an error,
```{r, error=TRUE}
remove(ints, 8)
```

but discard does not.
```{r}
discard(ints, 8) # ok
```

Discard and remove work also from the right.
```{r}
values(add(ints, 1:3))

values(discard(ints, 1))

values(discard(ints, 2, right=TRUE))
```

More details and examples are found in the online help (see `?Container`).


### Reference semantics and iterator
Being based on R6 classes, any `Container` object provides reference semantics.
```{r}
members <- print(container(c("Lisa", "Bob", "Joe")))

remove_Joe <- function(cont) discard(cont, "Joe")
remove_Joe(members)
members
```

```{r}
it <- iter(members)
print(it)

while(ithas_next(it)) {
    print(itget_next(it))
    print(it)
}
```

Once iterated to the last element, trying to iterate further leads to an error.
```{r, error=TRUE}
itget_next(it)
```


## Deque
```{r}
d <- deque(0L)
type(d)
d
```


#### addleft, count
```{r}
add(d, 1)
add(d, 2)
addleft(d, 1)
addleft(d, 2)
values(d)

count(d, 0)  # count number of 0s

count(d, 1)  # count number of 1s
```

#### peek, pop
A `peek` shows the last value, while `pop` shows and removes it afterwards. 
```{r}
peek(d)

pop(d)

pop(d)

d
```

Being a double-ended queue, both methods are also defined for the left side.
```{r}
peekleft(d)

popleft(d)

d
```

Invoking `peek` on an empty `Deque` gives `NULL` while `pop` stops with an error.
```{r, error=TRUE}
peek(deque())

pop(deque())
```

#### rotate, reverse
```{r}
values(add(d, rep(0, 3)))

values(rotate(d))    # rotate 1 to the right

values(rotate(d, 2))   # rotate 2 to the right

values(rotate(d, -3))  # rotate 3 to the left

values(addleft(d, 4:2))

values(reverse(d))
```

There is also a `+`-operator available, which does *not* work by reference but returns a fresh copy.
```{r}
d

5:6 + d

d + 5:6
```


#### Deque iterator
As a silly example, define a reverse perfect shuffler.
```{r}
reverse_ps <- function(x)
{
    it <- iter(seq_along(x))
    d <- deque(integer())
    
    while(ithas_next(it)) {
        itnext(it)
        d$add(itget(it))
        if (ithas_next(it)) d$addleft(itget_next(it))
    }
    x[values(d)]
}

(zz <- rep(c(0, 1), 10))

reverse_ps(zz)
```

## Set
```{r}
s1 <- set(1:3)
s1

add(s1, 1)  # does not change the set 
s1
```

#### set operators
```{r}
s1 <- set(c(1, 2,    4, 5))
s2 <- set(c(   2, 3,    5, 6))

s1 + s2

s1 / s2

s1 - s2

s1 < s2

s1 < (s1 + s2)

(s1 / s2) < s1

s1 == s2

s1 == s1

s1 > s2

(s1 + s2) > s2
```

## Dict
```{r}
(ages <- dict(c(Peter=24, Lisa=23, Bob=32)))

keys(ages)

peek(ages, "Lisa")

peek(ages, "Anna")
```

#### add, discard, remove
Due to the key-value semantic, several `Container ` methods are modified/extended to take the key argument.
```{r, error=TRUE}
values(add(ages, "Albert", 139))

add(ages, "Bob", 40)

has(ages, "Peter")

values(discard(ages, "Albert"))

# Trying to discard a non-existing key has no effect ...
discard(ages, "Albert")

# ... but trying to remove a non-existing key throws an error
remove(ages, "Albert")
```

#### add vs setval
Trying to set a value at a non-existing key throws an error unless the method is explicitly told to add it to the `Dict`.
```{r, error=TRUE}
setval(ages, "Anna", 23)
#ages[["Anna"]] <- 23             

setval(ages, "Anna", 23, add=TRUE)  # alternatively: add(ages, "Anna", 23)
#ages[["Anna", add=TRUE]] <- 23
ages
```

This allows fine control over the insert-behaviour of the `Dict`. If already existing, the value is overwritten.
```{r}
setval(ages, "Lisa", 11)
#ages[["Lisa"]] <- 11

values(ages)
```

#### peek, pop, and popitem
A similar control is provided via the different methods to retrieve elements.
```{r, error=TRUE}
pop(ages, "Lisa")

values(ages)

pop(ages, "Lisa")

getval(ages, "Lisa")

peek(ages, "Lisa")
```

Finally, the `Dict` could also be used as a sampler (without replacement).
```{r}
set.seed(123)
while(!ages$empty()) print(ages$popitem())
```

#### update
```{r}
shoplist <- dict(list(eggs=10, potatoes=10, bananas=5, apples=4))

shoplist2 <- dict(list(eggs=6, broccoli=4))

unlist(values(update(shoplist, shoplist2)))
```

#### dict operators
To enhance interactive use, various operators are provided for existing methods (see also overview table above). 
```{r, error=TRUE}
d <- dict()
d["A"] <- 1             # add(d, "A", 1)

d

d["A"] <- 2             # add(d, "A", 2)

d[["A"]] <- 2           # setval(d, "A", 2)

d[["B"]] <- 3           # setval(d, "A", 3)

d[["B", add=T]] <- 3    # setval(d, "A", 3, add=T)

d

d["A"]                  # peek(d, "A")

d[["C"]]                # getval(d, "C")

d["C"]                  # peek(d, "C")
```

```{r}
other <- dict(list(B=7, C=10))
d + other

d - other
```

## Conversion from and to base R types

```{r}
(co <- as.container(1:3))

as.vector(co)

as.list(co)

df <- data.frame(A=1:3, B=3:1)

d <- as.dict(df)
d

as.data.frame(d)
```

