---
title: "Manage parameter lists with dict"
author: "Roman Pahl"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Manage parameter lists with dict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Motivation

The original motivation for the development of this package actually was that
the author found himself writing countless checks and helper code over and over
again when managing parameter lists in larger applications.
It became apparent that something similar to python's
dictionary would make life easier and so the idea of the container
package was born.

The package has undergone some changes since it's initial version,
but the *dict* as a use-case for parameter lists remains very valid.
So without further ado, let's see how this works out in practice.

```{r}
library(container, warn.conflicts = FALSE)

# Define some parameters
params = dict(a = 1:10, b = "foo")
```


## Add or Replace
With a *dict* the problem of accidentally overriding an existing
parameter value is solved out of the box using the `add` function.

```{r, error = TRUE}
params = add(params, a = 0)

add(params, x = 0) # ok
```

Of course, it's also possible to indeed override a parameter.
```{r}
replace_at(params, a = 0)
```

What if you intend to replace something, but there is nothing to replace?
```{r, error = TRUE}
replace_at(params, x = 0)
```

Now you might wonder, what if *'I don't care if it is replaced or added'*.
That's easy.
```{r, error = TRUE}
replace_at(params, a = 0, .add=TRUE)

replace_at(params, x = 0, .add=TRUE)
```

That is, using `.add = TRUE` basically means, *'replace it, or, if it is not
there, just add it'*

Maybe you agree that even these simple use-cases already require some effort
when using base R lists.

## Extract
When extracting a parameter, you might want to be sure it exists and signal an
error otherwise.
```{r, error = TRUE}
at(params, "x")

at(params, "a", "b")
```

To extract a single raw element, use `at2`
```{r}
at2(params, "a")
```


Alternatively, you could use the standard access operators, which behave
like base R list and therefore return an empty *dict* (or `NULL`) if the
index is not found.
```{r}
params["x"]

params[["x"]]

params["a"]

params[["a"]]
```

### Default values
A nice property of the *dict* is that it provides an easy and flexible way to
manage default values.
```{r}
peek_at(params, "x")

peek_at(params, "x", .default = 3:1)
```

That is, if you *peek* at a non-existing parameter, by default an empty *dict*
is returned, but with the option to explicitly set the default. This also works
for multiple peeks.
```{r}
peek_at(params, "a", "x", "y", .default = 3:1)
```

## Remove
Similar to the above examples, the user can control how removal of
existing/non-existing parameters is handled. If you expect a parameter and
want to be signaled if it was not there, use `delete`.
```{r, error = TRUE}
delete_at(params, "x")

delete_at(params, "a") # ok
```

Otherwise to loosely delete a parameter, regardless of whether it
exists or not, use `discard`.
```{r}
discard_at(params, "a", "x")
```

It's important to note, that the "base R list way" to delete elements does not
work, because it just inserts a `NULL`.
```{r}
params[["a"]] <- NULL

params
```


## Merge
Last but not least, *dict* allows to easily merge and/or update parameter lists.
```{r}
par1 = dict(a = 1, b = "foo")
par2 = dict(b = "bar", x = 2, y = 3)

update(par1, par2)
```

As can be seen, existing parameters are updated and new parameters added.
Using `as.dict` you can also do this with ordinary lists.
```{r}
update(par1, as.dict(list(b = "my b", x = 100)))
```

That's it. I hope, it will free you some time and safe some bugs next time you
need to manage parameter lists.

As a very last note, keep in mind that since container version 1.0.0, *dict*
elements are always sorted by their name, while you are still able to
access elements by position (based on the sorted values).
```{r}
d = dict(x = 1, z = 2, a = 3)
d

d[[1]]

d[2:3]
```





